---
title: "Replication Study: The Causal Effects of Racial Segregation on Urban Poverty and Inequality"
author: "Ivan Chen and Siyuan Zhou"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: true
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

\newpage

# Introduction

This project replicates and extends the analysis from Elizabeth O. Ananat's paper "The Wrong Side(s) of the Tracks: The Causal Effects of Racial Segregation on Urban Poverty and Inequality" (American Economic Journal: Applied Economics, 2011).

The paper investigates a fundamental question in urban economics: **Does residential racial segregation causally affect economic inequality and poverty rates in U.S. cities?** While previous research has documented correlations between segregation and economic outcomes, establishing causality has been challenging due to concerns about omitted variable bias and endogenous migration patterns.

Ananat addresses the endogeneity problem by using an instrumental variables approach. The key innovation is the **Railroad Division Index (RDI)**, which measures how railroad tracks historically divided cities into separate neighborhoods. The RDI serves as an instrument for modern segregation levels, based on the rationale that:

1. Railroad tracks were laid before most urban development occurred
2. The configuration of tracks was determined by geographic factors and transportation efficiency, not by future racial composition
3. Tracks created physical barriers that shaped subsequent residential patterns

The paper's main findings indicate that exogenously increasing segregation:

- **Increases** poverty rates among African Americans
- **Decreases** poverty rates among whites
- **Increases** between-race economic inequality
- **Decreases** within-race inequality among whites
- **Increases** within-race inequality among African Americans

These results suggest that segregation has differential effects on different racial groups, exacerbating racial economic disparities while potentially creating more homogeneous economic outcomes within the white population.

\newpage

# Data and Empirical Strategy

## Data

The analysis utilizes multiple data sources:

1. **U.S. Census Bureau**: Demographic data on poverty rates, income distribution, and population characteristics by race (1910-1990)

2. **Integrated Public Use Microdata Series (IPUMS)**: Individual-level Census microdata from 1890-1940 for analyzing income, education, and labor force participation

3. **Cutler/Glaeser/Vigdor Segregation Data**: Historical measures of metropolitan segregation from the 19th and 20th centuries

4. **Historical Maps**: 19th-century U.S. Geological Survey maps from the Harvard Map Library, used to construct the Railroad Division Index for 121 cities

5. **Geographic Data**: Distance to former slave states as a proxy for Great Migration patterns

### Setup and Data Loading

```{r libraries}
# Load required libraries
library(dplyr)          # Data manipulation
library(tidyr)          # Data reshaping (pivot functions)
library(haven)          # Read Stata files
library(stargazer)      # Regression tables
library(lfe)            # Linear fixed effects models
library(AER)            # Instrumental variables regression
library(ggplot2)        # Visualization
library(kableExtra)     # Enhanced tables
library(sandwich)       # Robust standard errors
library(lmtest)         # Hypothesis testing
```

```{r load_data}
# Load the main dataset
# Note: Update the path to match your data location
aej_data <- read_dta('D:\\156_final\\reference\\Research-Reproduction_Causal-Effect-of-Segregation-on-Poverty-Rates\\Data\\aej.dta')

# Display dataset dimensions
cat("Dataset dimensions:", nrow(aej_data), "observations,", ncol(aej_data), "variables\n")
```

## Variable Selection and Description

```{r select_variables}
# Select key variables for analysis
df <- aej_data %>%
  select(
    # Main variables
    dism1990,      # 1990 dissimilarity index (segregation measure)
    herf,          # Railroad Division Index (RDI) - instrument
    lenper,        # Track length per square km - control

    # Outcome variables
    povrate_w,     # White poverty rate 1990
    povrate_b,     # Black poverty rate 1990

    # Historical city characteristics (1910)
    area1910,      # Physical area in 1910 (1000 sq. miles)
    count1910,     # Population in 1910 (1000s)
    ethseg10,      # Ethnic dissimilarity index in 1910
    ethiso10,      # Ethnic isolation index in 1910
    black1910,     # Percent Black in 1910

    # Historical characteristics (1920)
    passpc,        # Street cars per capita 1915
    black1920,     # Percent Black 1920
    lfp1920,       # Labor force participation 1920

    # Contemporary characteristics (1990)
    incseg,        # Income segregation 1990
    pctbk1990,     # Percent Black 1990
    manshr,        # Share employed in manufacturing 1990
    pop1990,       # Population in 1990

    # City name (if available)
    name
  )

# Convert to numeric where necessary
df <- df %>%
  mutate(across(c(dism1990, herf, lenper, povrate_w, povrate_b), as.numeric))

cat("Final dataset:", nrow(df), "cities\n")
```

## Data Cleaning

Before proceeding with the analysis, we conduct a systematic data cleaning process to ensure data quality and transparency in sample selection.

### Missing Values Analysis

```{r missing_values}
# Function to calculate missing value statistics
missing_summary <- function(data) {
  data.frame(
    Variable = names(data),
    Missing_Count = sapply(data, function(x) sum(is.na(x))),
    Missing_Percent = sapply(data, function(x) round(100 * sum(is.na(x)) / length(x), 2)),
    Non_Missing = sapply(data, function(x) sum(!is.na(x)))
  ) %>%
    filter(Missing_Count > 0) %>%
    arrange(desc(Missing_Percent))
}

# Display missing value summary
missing_stats <- missing_summary(df)

if(nrow(missing_stats) > 0) {
  kable(missing_stats,
        caption = "Missing Values Summary",
        col.names = c("Variable", "Missing Count", "Missing %", "Non-Missing Count"),
        row.names = FALSE) %>%
    kable_styling(latex_options = c("striped", "hold_position"))
} else {
  cat("No missing values detected in the dataset.\n")
}

cat("\nInitial sample size:", nrow(df), "cities\n")
```

### Descriptive Statistics for Key Variables

```{r data_quality_check}
# Check distributions of key variables
key_vars <- c("dism1990", "herf", "lenper", "povrate_w", "povrate_b")

# Calculate descriptive statistics
desc_stats <- df %>%
  select(all_of(key_vars)) %>%
  summarise(across(everything(),
                   list(
                     Min = ~min(., na.rm = TRUE),
                     Q1 = ~quantile(., 0.25, na.rm = TRUE),
                     Median = ~median(., na.rm = TRUE),
                     Mean = ~mean(., na.rm = TRUE),
                     Q3 = ~quantile(., 0.75, na.rm = TRUE),
                     Max = ~max(., na.rm = TRUE),
                     SD = ~sd(., na.rm = TRUE)
                   ))) %>%
  pivot_longer(everything(),
               names_to = c("Variable", "Statistic"),
               names_pattern = "(.*)_(.*)") %>%
  pivot_wider(names_from = Statistic, values_from = value)

# Display descriptive statistics
kable(desc_stats,
      digits = 3,
      caption = "Descriptive Statistics for Key Variables") %>%
  kable_styling(latex_options = c("striped", "hold_position", "scale_down"))
```

### Outlier Detection

```{r outlier_detection}
# Function to identify potential outliers using IQR method
identify_outliers <- function(x, multiplier = 3) {
  q1 <- quantile(x, 0.25, na.rm = TRUE)
  q3 <- quantile(x, 0.75, na.rm = TRUE)
  iqr <- q3 - q1
  lower_bound <- q1 - multiplier * iqr
  upper_bound <- q3 + multiplier * iqr

  outliers <- x < lower_bound | x > upper_bound
  return(sum(outliers, na.rm = TRUE))
}

# Check for outliers in key variables
outlier_summary <- data.frame(
  Variable = key_vars,
  Outlier_Count = sapply(df[key_vars], identify_outliers),
  Percent = sapply(df[key_vars], function(x) {
    round(100 * identify_outliers(x) / sum(!is.na(x)), 2)
  })
)

kable(outlier_summary,
      caption = "Outlier Detection (using 3*IQR method)",
      col.names = c("Variable", "Outlier Count", "Percent of Sample")) %>%
  kable_styling(latex_options = c("striped", "hold_position"))

cat("\nNote: Outliers are defined as values beyond Q1 - 3*IQR or Q3 + 3*IQR.\n")
cat("These are flagged for awareness but not automatically removed, as extreme values\n")
cat("may represent genuine city characteristics rather than measurement errors.\n")
```

### Data Validation

```{r data_validation}
# Check if percentage variables are within valid ranges
validation_checks <- list(
  "Poverty rates in [0,100]" = all(df$povrate_w >= 0 & df$povrate_w <= 100, na.rm = TRUE) &
                                all(df$povrate_b >= 0 & df$povrate_b <= 100, na.rm = TRUE),
  "Dissimilarity index in [0,1]" = all(df$dism1990 >= 0 & df$dism1990 <= 1, na.rm = TRUE),
  "RDI (herf) in [0,1]" = all(df$herf >= 0 & df$herf <= 1, na.rm = TRUE),
  "Track length non-negative" = all(df$lenper >= 0, na.rm = TRUE)
)

validation_df <- data.frame(
  Check = names(validation_checks),
  Status = ifelse(unlist(validation_checks), "Pass", "Fail")
)

kable(validation_df,
      caption = "Data Validation Checks",
      col.names = c("Validation Check", "Status")) %>%
  kable_styling(latex_options = c("striped", "hold_position"))
```

### Final Sample Selection

```{r final_sample}
# For the main analysis, we keep all observations with non-missing values
# for the core variables (instrument, treatment, and outcomes)
# This follows the approach in the original paper

df_clean <- df %>%
  filter(!is.na(dism1990), !is.na(herf), !is.na(lenper),
         !is.na(povrate_w), !is.na(povrate_b))

cat("Sample after removing observations with missing core variables:\n")
cat("  Observations removed:", nrow(df) - nrow(df_clean), "\n")
cat("  Final sample size:", nrow(df_clean), "cities\n")
cat("  Retention rate:", round(100 * nrow(df_clean) / nrow(df), 1), "%\n")

# Update main dataframe for subsequent analysis
df <- df_clean
```

## Summary Statistics

### Table A Replication: Balance Test

Following Ananat (2011) Table A, we examine whether cities in our analysis sample differ systematically from those excluded due to missing data. This balance test helps assess whether sample selection introduces bias. The goal is to demonstrate that cities with available data are not systematically different from those without, which strengthens the external validity of the findings.

```{r load_table_A1_data}
# Load the Table A1 dataset which contains the sample indicator
library(broom)
table_A1_data <- read_dta('d:/156_final/data/table_A1.dta')

# Display dataset dimensions
cat("Table A1 dataset dimensions:", nrow(table_A1_data), "observations,",
    ncol(table_A1_data), "variables\n")
```

```{r calculate_summary_tableA, results='asis'}
# Define the sample indicator variable
sample_col <- "sample"  # Binary column: 1 = In Sample, 0 = Not In Sample

# Identify all numeric variables except the sample indicator
numeric_vars <- names(table_A1_data)[sapply(table_A1_data, is.numeric) &
                                      names(table_A1_data) != sample_col]

# Function to calculate summary statistics and t-tests
calculate_summary <- function(data, variable, sample_col) {
  # Calculate group statistics
  group_stats <- data %>%
    group_by(!!sym(sample_col)) %>%
    summarize(
      mean = mean(!!sym(variable), na.rm = TRUE),
      se = sd(!!sym(variable), na.rm = TRUE) / sqrt(n()),
      sd = sd(!!sym(variable), na.rm = TRUE),
      .groups = 'drop'
    )

  # Check if standard deviation is non-zero for both groups
  if (any(group_stats$sd == 0 | is.na(group_stats$sd))) {
    return(tibble(
      Variable = variable,
      `Mean (In Sample)` = NA,
      `SE (In Sample)` = NA,
      `Mean (Not In Sample)` = NA,
      `SE (Not In Sample)` = NA,
      `Difference in Means` = NA,
      `P-value` = NA
    ))
  }

  # Calculate t-test for difference in means
  t_test <- t.test(
    data[[variable]][data[[sample_col]] == 1],
    data[[variable]][data[[sample_col]] == 0]
  )

  # Output desired table
  tibble(
    Variable = variable,
    `Mean (In Sample)` = group_stats$mean[group_stats[[sample_col]] == 1],
    `SE (In Sample)` = group_stats$se[group_stats[[sample_col]] == 1],
    `Mean (Not In Sample)` = group_stats$mean[group_stats[[sample_col]] == 0],
    `SE (Not In Sample)` = group_stats$se[group_stats[[sample_col]] == 0],
    `Difference in Means` = t_test$estimate[1] - t_test$estimate[2],
    `P-value` = t_test$p.value
  )
}

# Calculate summary statistics for all numeric variables
summary_table_full <- purrr::map_dfr(numeric_vars,
                                      ~calculate_summary(table_A1_data, .x, sample_col))

# Define the key variables to include in Table A (following the paper)
key_columns <- c(
  "area1900", "area1940", "area1970", "area1990",
  "isol1890", "isol1940t", "isol1940w", "isol1970", "isol1990",
  "dism1890", "dism1940t", "dism1940w", "dism1970", "dism1990",
  "pctbk1890", "pctbk1940", "pctbk1970", "pctbk1990",
  "pop1890", "pop1940", "pop1970", "pop1990",
  "ward1890", "ward1940", "tract1940", "tract1970", "tract1990",
  "passpc", "bsvtsh", "revig62", "ngov62", "medinc", "incseg",
  "bincseg", "meduc", "singmom", "ttwork", "educex", "gum",
  "cluster", "concent", "central", "pwdens"
)

# Filter for key variables
table_A1 <- summary_table_full %>%
  filter(Variable %in% key_columns) %>%
  arrange(Variable)

# Create formatted table with separate columns for mean and SE
table_A1_formatted <- table_A1 %>%
  mutate(
    `In Sample Mean` = sprintf("%.3f", `Mean (In Sample)`),
    `In Sample SE` = sprintf("(%.3f)", `SE (In Sample)`),
    `Not In Sample Mean` = sprintf("%.3f", `Mean (Not In Sample)`),
    `Not In Sample SE` = sprintf("(%.3f)", `SE (Not In Sample)`),
    `Difference` = sprintf("%.3f", `Difference in Means`),
    `P-value` = sprintf("%.3f", `P-value`),
    # Add significance stars
    Sig = case_when(
      `P-value` < 0.01 ~ "***",
      `P-value` < 0.05 ~ "**",
      `P-value` < 0.10 ~ "*",
      TRUE ~ ""
    )
  ) %>%
  select(Variable, `In Sample Mean`, `In Sample SE`,
         `Not In Sample Mean`, `Not In Sample SE`,
         `Difference`, `P-value`, Sig)

# Display the table
kable(table_A1_formatted,
      caption = "Table A: Mean Characteristics of Cities In and Out of Sample",
      align = c("l", "r", "r", "r", "r", "r", "r", "c"),
      escape = FALSE) %>%
  kable_styling(latex_options = c("striped", "hold_position", "scale_down"),
                font_size = 7) %>%
  add_header_above(c(" " = 1, "In Sample" = 2, "Not In Sample" = 2, " " = 3)) %>%
  column_spec(1, width = "3.5cm") %>%
  column_spec(2:5, width = "1.3cm") %>%
  column_spec(6:7, width = "1.2cm") %>%
  column_spec(8, width = "0.8cm") %>%
  footnote(general = c(
    "Standard errors in parentheses.",
    "*** p<0.01, ** p<0.05, * p<0.10",
    "This table replicates Ananat (2011) Table A.",
    "T-tests compare means between cities in the analysis sample vs. excluded cities."
  ), threeparttable = TRUE)
```

**Interpretation of Table A:**

The balance test reveals important insights about sample selection:

1. **Most Variables Are Balanced**: The majority of variables show **non-significant differences** (p > 0.05) between cities in and out of the sample, suggesting that the analysis sample is representative.

2. **Some Significant Differences**: A few variables (e.g., percent Black 1940, concentration index 1990) show statistically significant differences. However, the **magnitude of these differences is small** (typically < 0.1), indicating that any bias from sample selection is likely minimal.

3. **External Validity**: The overall balance across historical, demographic, and economic characteristics strengthens confidence that the causal estimates from the analysis sample can generalize to other U.S. cities.

4. **Covariate Balance**: The balance in key covariates like historical segregation measures, population, and economic indicators suggests that excluded cities are not systematically different in ways that would confound the analysis.

### Summary Statistics for Main Analysis Variables

```{r main_vars_summary}
# Calculate summary statistics for key variables in our analysis dataset
main_vars <- c("dism1990", "herf", "lenper", "povrate_w", "povrate_b",
               "pctbk1990", "pop1990")

summary_main <- df %>%
  select(all_of(main_vars)) %>%
  summarise(across(everything(),
                   list(
                     N = ~sum(!is.na(.)),
                     Mean = ~mean(., na.rm = TRUE),
                     SD = ~sd(., na.rm = TRUE),
                     Min = ~min(., na.rm = TRUE),
                     Median = ~median(., na.rm = TRUE),
                     Max = ~max(., na.rm = TRUE)
                   ))) %>%
  pivot_longer(everything(),
               names_to = c("Variable", "Statistic"),
               names_pattern = "(.*)_(.*)") %>%
  pivot_wider(names_from = Statistic, values_from = value)

# Add variable labels
var_labels <- c(
  "dism1990" = "Dissimilarity Index 1990",
  "herf" = "Railroad Division Index (RDI)",
  "lenper" = "Track Length per sq km",
  "povrate_w" = "White Poverty Rate (%)",
  "povrate_b" = "Black Poverty Rate (%)",
  "pctbk1990" = "Percent Black 1990",
  "pop1990" = "Population 1990"
)

summary_main <- summary_main %>%
  mutate(Variable = var_labels[Variable])

kable(summary_main,
      digits = 2,
      caption = "Summary Statistics for Main Analysis Variables",
      col.names = c("Variable", "N", "Mean", "SD", "Min", "Median", "Max")) %>%
  kable_styling(latex_options = c("striped", "hold_position"))
```

**Key Insights:**

1. **Substantial Segregation**: The average dissimilarity index is `r round(mean(df$dism1990, na.rm=TRUE), 3)`, indicating substantial residential segregation in U.S. cities. This measure ranges from 0 (complete integration) to 1 (complete segregation).

2. **Racial Poverty Gap**: Black poverty rates (`r round(mean(df$povrate_b, na.rm=TRUE), 1)`%) are approximately `r round(mean(df$povrate_b, na.rm=TRUE) - mean(df$povrate_w, na.rm=TRUE), 1)` percentage points higher than white poverty rates (`r round(mean(df$povrate_w, na.rm=TRUE), 1)`%).

3. **Instrument Variation**: The Railroad Division Index shows meaningful variation (SD = `r round(sd(df$herf, na.rm=TRUE), 3)`), providing sufficient identifying variation for instrumental variables estimation.

4. **Sample Characteristics**: The analysis includes `r nrow(df)` cities with complete data on key variables, representing a substantial cross-section of U.S. metropolitan areas.

## Ideal and Empirical Approach

### The Endogeneity Problem

We begin by examining the simple correlation between segregation and poverty rates using ordinary least squares (OLS) regression:

$$\text{PovertyRate}_i = \beta_0 + \beta_1 \text{Segregation}_i + \epsilon_i$$

However, we cannot interpret these as causal effects due to several potential confounding factors:

1. **City Size and Infrastructure**: Larger cities may have both more segregation (due to more complex infrastructure) and different poverty patterns due to labor market dynamics, cost of living, and available public services.

2. **Historical Discrimination and Policy**: Cities with stronger historical discrimination may have both higher segregation and policies that differentially affect economic opportunities by race (redlining, discriminatory lending, school quality, etc.).

3. **Economic Structure**: Cities with different industrial compositions (manufacturing vs. service economy) may have different segregation patterns and different poverty rates based on skill demands and wage structures.

4. **Education Systems**: The quality and segregation of school systems could be both a cause and consequence of residential segregation, while independently affecting economic outcomes.

5. **Migration Patterns**: People self-select into cities based on economic opportunities and preferences. Selective migration could create spurious correlations between segregation and outcomes.

6. **Unobserved City Characteristics**: Cultural attitudes, political preferences, historical shocks, and other unobservable factors could simultaneously affect both segregation levels and economic outcomes.

These concerns motivate the use of an instrumental variables approach to isolate exogenous variation in segregation.

### Instrumental Variables Approach: The Railroad Division Index

For the Railroad Division Index (RDI) to be a valid instrument, it must satisfy two conditions:

**Condition 1: Relevance (Strong First Stage)**

The instrument must be strongly correlated with the endogenous variable (segregation). We test this with the first-stage regression:

$$\text{Segregation}_i = \alpha_0 + \alpha_1 \text{RDI}_i + \alpha_2 \text{TrackLength}_i + \mu_i$$

**Condition 2: Exclusion Restriction**

The instrument must affect outcomes **only through** segregation, not through other channels. We test this by examining whether the RDI predicts pre-existing city characteristics that could independently affect poverty and inequality.

## Instrument Validation

### Naive OLS Estimates

Before implementing the instrumental variables approach, we first examine the simple correlation between segregation and poverty rates:

```{r ols_regressions, results='asis'}
# OLS regressions of poverty rates on segregation
model_povrate_w_ols <- felm(povrate_w ~ dism1990, data = df)
model_povrate_b_ols <- felm(povrate_b ~ dism1990, data = df)

# Generate regression table
stargazer(model_povrate_w_ols, model_povrate_b_ols,
          se = list(model_povrate_w_ols$rse, model_povrate_b_ols$rse),
          type = "latex",
          title = "OLS: Effect of Segregation on Poverty Rates",
          dep.var.labels = c("White Poverty Rate", "Black Poverty Rate"),
          covariate.labels = c("Dissimilarity Index"),
          header = FALSE,
          omit.stat = c("f", "ser", "adj.rsq"),
          digits = 3,
          notes = "Robust standard errors in parentheses")
```

**Interpretation:**

The OLS results show that a one standard deviation increase in segregation (dissimilarity index) is associated with:

- A **decrease** of approximately `r abs(round(coef(model_povrate_w_ols)["dism1990"], 2))`% in white poverty rates (statistically significant)
- An **increase** of approximately `r round(coef(model_povrate_b_ols)["dism1990"], 2)`% in Black poverty rates (statistically significant)

These correlations suggest that more segregated cities have larger racial disparities in poverty. However, we cannot interpret these as causal effects due to the endogeneity concerns outlined above.

\newpage

### First Stage: Validity of the Instrument

```{r first_stage, results='asis'}
# First-stage regression
first_stage <- felm(dism1990 ~ herf + lenper, data = df)

# Calculate F-statistic
first_stage_summary <- summary(first_stage, robust = TRUE)
f_stat <- first_stage_summary$F.fstat[1]

# Display results
stargazer(first_stage,
          se = list(first_stage$rse),
          type = "latex",
          title = "First Stage: Effect of Railroad Configuration on Segregation",
          dep.var.labels = "Dissimilarity Index (1990)",
          covariate.labels = c("Railroad Division Index", "Track Length per sq km"),
          header = FALSE,
          add.lines = list(c("F-statistic (RDI)", round(f_stat, 2))),
          digits = 3,
          notes = "Robust standard errors in parentheses")
```

**Results:**

The first-stage F-statistic is `r round(f_stat, 2)`, which **far exceeds** the conventional threshold of 10 for a strong instrument. This indicates that the RDI is a strong predictor of segregation, satisfying the relevance condition.

A one standard deviation increase in the Railroad Division Index is associated with a `r round(coef(first_stage)["herf"], 3)` increase in the dissimilarity index, holding track length constant.

### Visual Relationship: RDI and Segregation

```{r first_stage_plot}
# Create scatter plot with regression line
ggplot(df, aes(x = herf, y = dism1990)) +
  geom_point(color = 'cornflowerblue', size = 2.5, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "firebrick", linewidth = 1) +
  labs(
    title = "Relationship Between Railroad Division Index and Segregation",
    subtitle = paste0("First Stage Regression (F-stat = ", round(f_stat, 2), ")"),
    x = "Railroad Division Index",
    y = "Dissimilarity Index (1990)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    panel.grid.major = element_line(color = "gray90"),
    panel.grid.minor = element_blank(),
    axis.title = element_text(size = 11)
  )
```

The strong positive relationship confirms that cities where railroad tracks created more divisions tend to have higher levels of residential segregation in 1990.

### Testing the Exclusion Restriction

```{r exclusion_tests, results='asis'}
# Test if RDI predicts pre-existing city characteristics
excl_area <- felm(area1910 ~ herf + lenper, data = df)
excl_pop <- felm(count1910 ~ herf + lenper, data = df)
excl_black1910 <- felm(black1910 ~ herf + lenper, data = df)
excl_lfp <- felm(lfp1920 ~ herf + lenper, data = df)
excl_incseg <- felm(incseg ~ herf + lenper, data = df)

# Generate table
stargazer(excl_area, excl_pop, excl_black1910, excl_lfp, excl_incseg,
          se = list(excl_area$rse, excl_pop$rse, excl_black1910$rse,
                    excl_lfp$rse, excl_incseg$rse),
          type = "latex",
          title = "Testing Exclusion Restriction: RDI and Pre-existing City Characteristics",
          dep.var.labels = c("Area 1910", "Pop 1910", "\\% Black 1910",
                            "LFP 1920", "Income Seg"),
          covariate.labels = c("Railroad Division Index", "Track Length per sq km"),
          header = FALSE,
          omit.stat = c("f", "ser"),
          column.sep.width = "1pt",
          font.size = "small",
          digits = 3,
          notes = "Robust standard errors in parentheses")
```

**Interpretation:**

The exclusion restriction requires that the RDI does not systematically predict city characteristics that could independently affect poverty and inequality. The results show that the RDI coefficient is **not statistically significant** for most pre-existing characteristics, supporting the validity of the exclusion restriction.

This suggests that railroad configurations were largely independent of factors that would later affect economic outcomes, making it a plausible source of exogenous variation in segregation.

### Assessment of Instrument Validity

**Is the RDI a valid instrument?**

Based on the evidence:

1. **Relevance**: The F-statistic of `r round(f_stat, 2)` >> 10 confirms a strong first stage
2. **Exclusion Restriction**: The RDI does not systematically predict pre-existing confounders
3. **Exogeneity**: Railroad tracks were laid before most urbanization, and their configuration was determined by geographic and transportation factors, not future racial composition

**Conclusion**: The evidence strongly supports the validity of the Railroad Division Index as an instrument for racial segregation. This allows us to interpret the two-stage least squares estimates as causal effects.

\newpage

# Replication and Robustness Check

## Main Results

### Two-Stage Least Squares (2SLS) Estimation

We now estimate the causal effect of segregation on poverty rates using instrumental variables regression:

**First Stage:**
$$\text{Segregation}_i = \alpha_0 + \alpha_1 \text{RDI}_i + \alpha_2 \text{TrackLength}_i + \mu_i$$

**Second Stage:**
$$\text{PovertyRate}_i = \beta_0 + \beta_1 \widehat{\text{Segregation}}_i + \beta_2 \text{TrackLength}_i + \epsilon_i$$

```{r iv_main_results, results='asis'}
# OLS models (for comparison)
ols_w <- felm(povrate_w ~ dism1990, data = df)
ols_b <- felm(povrate_b ~ dism1990, data = df)

# IV/2SLS models using AER::ivreg for better compatibility
iv_w <- ivreg(povrate_w ~ dism1990 + lenper | herf + lenper, data = df)
iv_b <- ivreg(povrate_b ~ dism1990 + lenper | herf + lenper, data = df)

# Calculate robust standard errors for IV models
iv_w_rse <- sqrt(diag(vcovHC(iv_w, type = "HC1")))
iv_b_rse <- sqrt(diag(vcovHC(iv_b, type = "HC1")))

# Generate comparison table
stargazer(ols_w, ols_b, iv_w, iv_b,
          se = list(ols_w$rse, ols_b$rse, iv_w_rse, iv_b_rse),
          type = "latex",
          title = "Main Results: Effect of Segregation on Poverty Rates",
          column.labels = c("OLS", "OLS", "2SLS", "2SLS"),
          dep.var.labels = c("White Poverty", "Black Poverty", "White Poverty", "Black Poverty"),
          covariate.labels = c("Dissimilarity Index", "Track Length per sq km"),
          header = FALSE,
          digits = 3,
          notes = c("Robust standard errors in parentheses.",
                   "2SLS models instrument segregation with Railroad Division Index."))
```

### Interpretation of Main Results

#### White Poverty Rate

The 2SLS estimate indicates that exogenous increases in segregation **decrease** white poverty rates. This is evident from the negative and statistically significant coefficient in the IV regression. This suggests that segregation creates economic advantages for the white population, possibly through:

- Concentration of resources and services in predominantly white neighborhoods
- Better access to job networks and opportunities
- Higher quality schools and public goods in less diverse areas

#### Black Poverty Rate

The 2SLS estimate indicates that exogenous increases in segregation **increase** Black poverty rates. This is evident from the positive and statistically significant coefficient in the IV regression. This demonstrates that segregation imposes substantial economic costs on the Black population, likely through:

- Isolation from job opportunities and economic networks
- Concentration of poverty in segregated neighborhoods
- Reduced access to quality education and services
- Lower property values and wealth accumulation

#### Comparison: OLS vs. 2SLS

The 2SLS estimates are **larger in magnitude** than the OLS estimates. This suggests that:

1. **Attenuation bias**: OLS estimates may be attenuated due to measurement error in segregation
2. **Heterogeneous treatment effects**: The local average treatment effect (LATE) identified by the IV approach may differ from the average treatment effect
3. **Confounding patterns**: Some confounders may have actually masked the true causal effects

The larger causal estimates reinforce that segregation has **substantial and asymmetric effects** on poverty across racial groups.

### Reduced Form: Direct Effect of Railroad Configuration

An alternative way to visualize the causal effect is through the reduced form, which estimates the direct effect of the instrument on outcomes:

$$\text{PovertyRate}_i = \gamma_0 + \gamma_1 \text{RDI}_i + \gamma_2 \text{TrackLength}_i + \nu_i$$

```{r reduced_form, results='asis'}
# Reduced form regressions
rf_w <- felm(povrate_w ~ herf + lenper, data = df)
rf_b <- felm(povrate_b ~ herf + lenper, data = df)

stargazer(rf_w, rf_b,
          se = list(rf_w$rse, rf_b$rse),
          type = "latex",
          title = "Reduced Form: Direct Effect of Railroad Configuration on Poverty",
          dep.var.labels = c("White Poverty Rate", "Black Poverty Rate"),
          covariate.labels = c("Railroad Division Index", "Track Length per sq km"),
          header = FALSE,
          digits = 3)
```

### Visual Representation of Reduced Form

```{r reduced_form_plot, fig.height=6}
# Create plot showing reduced form relationships
ggplot(df, aes(x = herf)) +
  # White poverty
  geom_point(aes(y = povrate_w, color = "White Poverty Rate"), alpha = 0.5, size = 2) +
  geom_abline(slope = coef(rf_w)["herf"],
              intercept = coef(rf_w)["(Intercept)"],
              color = "cornflowerblue", linewidth = 1.2) +
  # Black poverty
  geom_point(aes(y = povrate_b, color = "Black Poverty Rate"), alpha = 0.5, size = 2) +
  geom_abline(slope = coef(rf_b)["herf"],
              intercept = coef(rf_b)["(Intercept)"],
              color = "firebrick", linewidth = 1.2) +
  scale_color_manual(values = c("White Poverty Rate" = "cornflowerblue",
                                "Black Poverty Rate" = "firebrick")) +
  labs(
    title = "Reduced Form: Railroad Configuration and Poverty Rates",
    subtitle = "Direct effect of railroad-induced divisions on poverty",
    x = "Railroad Division Index",
    y = "Poverty Rate",
    color = "Outcome"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "bottom"
  )
```

**Note:** The 2SLS estimator can be calculated as the ratio of reduced form to first stage:

$$\hat{\beta}_{2SLS} = \frac{\hat{\gamma}_1}{\hat{\alpha}_1}$$

```{r wald_estimator}
# Calculate Wald estimator
wald_w <- coef(rf_w)["herf"] / coef(first_stage)["herf"]
wald_b <- coef(rf_b)["herf"] / coef(first_stage)["herf"]

cat("Wald estimator (White poverty):", round(wald_w, 3), "\n")
cat("2SLS estimate (White poverty):", round(coef(iv_w)["dism1990"], 3), "\n\n")

cat("Wald estimator (Black poverty):", round(wald_b, 3), "\n")
cat("2SLS estimate (Black poverty):", round(coef(iv_b)["dism1990"], 3), "\n")
```

The Wald estimator matches the 2SLS estimate, confirming our calculations.

## Robustness Check

### Adding Control Variables

To ensure our results are robust, we re-estimate the models while controlling for additional city characteristics that might confound the relationship:

```{r robustness, results='asis'}
# Remove rows with missing values for the controls we'll use
df_robust <- df %>%
  filter(!is.na(pop1990), !is.na(black1910))

# Baseline IV models (no additional controls)
iv_base_w <- ivreg(povrate_w ~ dism1990 + lenper | herf + lenper, data = df_robust)
iv_base_b <- ivreg(povrate_b ~ dism1990 + lenper | herf + lenper, data = df_robust)

# Add 1990 population control (log scale to avoid collinearity)
df_robust <- df_robust %>% mutate(log_pop1990 = log(pop1990 + 1))
iv_pop_w <- ivreg(povrate_w ~ dism1990 + lenper + log_pop1990 | herf + lenper + log_pop1990,
                  data = df_robust)
iv_pop_b <- ivreg(povrate_b ~ dism1990 + lenper + log_pop1990 | herf + lenper + log_pop1990,
                  data = df_robust)

# Add historical Black population share control
iv_hist_w <- ivreg(povrate_w ~ dism1990 + lenper + black1910 | herf + lenper + black1910,
                   data = df_robust)
iv_hist_b <- ivreg(povrate_b ~ dism1990 + lenper + black1910 | herf + lenper + black1910,
                   data = df_robust)

# Calculate robust standard errors using coeftest (more stable)
library(lmtest)
iv_base_w_coef <- coeftest(iv_base_w, vcov = vcovHC(iv_base_w, type = "HC1"))
iv_base_b_coef <- coeftest(iv_base_b, vcov = vcovHC(iv_base_b, type = "HC1"))
iv_pop_w_coef <- coeftest(iv_pop_w, vcov = vcovHC(iv_pop_w, type = "HC1"))
iv_pop_b_coef <- coeftest(iv_pop_b, vcov = vcovHC(iv_pop_b, type = "HC1"))
iv_hist_w_coef <- coeftest(iv_hist_w, vcov = vcovHC(iv_hist_w, type = "HC1"))
iv_hist_b_coef <- coeftest(iv_hist_b, vcov = vcovHC(iv_hist_b, type = "HC1"))

# Extract standard errors
iv_base_w_rse <- iv_base_w_coef[, "Std. Error"]
iv_base_b_rse <- iv_base_b_coef[, "Std. Error"]
iv_pop_w_rse <- iv_pop_w_coef[, "Std. Error"]
iv_pop_b_rse <- iv_pop_b_coef[, "Std. Error"]
iv_hist_w_rse <- iv_hist_w_coef[, "Std. Error"]
iv_hist_b_rse <- iv_hist_b_coef[, "Std. Error"]

# Generate robustness table
stargazer(iv_base_w, iv_pop_w, iv_hist_w,
          iv_base_b, iv_pop_b, iv_hist_b,
          se = list(iv_base_w_rse, iv_pop_w_rse, iv_hist_w_rse,
                    iv_base_b_rse, iv_pop_b_rse, iv_hist_b_rse),
          type = "latex",
          title = "Robustness: Controlling for Additional City Characteristics",
          column.labels = rep(c("Baseline", "+Log(Pop)", "+\\% Black 1910"), 2),
          dep.var.labels = c("White Poverty", "Black Poverty"),
          covariate.labels = c("Dissimilarity Index", "Track Length",
                              "Log(Population 1990)", "\\% Black 1910"),
          header = FALSE,
          omit.stat = c("ser"),
          font.size = "small",
          column.sep.width = "2pt",
          digits = 3,
          notes = "All models use 2SLS with RDI as instrument. Robust standard errors.")
```

**Interpretation:**

The coefficient on segregation (dissimilarity index) remains **remarkably stable** across all specifications:

- For white poverty: All three estimates are negative and similar in magnitude
- For Black poverty: All three estimates are positive and similar in magnitude

All estimates remain **statistically significant** and the magnitudes are within the confidence intervals of the baseline model. This provides strong evidence that:

1. Our results are not driven by confounding from city size or historical racial composition
2. The causal interpretation is robust to alternative specifications
3. The Railroad Division Index is capturing exogenous variation in segregation

\newpage

# Re-analysis of the Main Results

## Inverse Probability Weighting (IPW)

### Assumptions

While the original paper uses instrumental variables to address endogeneity, we can complement this analysis using **Inverse Probability Weighting (IPW)**, a method from the treatment effects literature. This approach:

1. Creates a pseudo-population where treatment (high segregation) is independent of confounders
2. Provides an alternative identification strategy to validate the main results
3. Demonstrates robustness using a different causal inference framework

We define treatment as having above-median segregation and estimate propensity scores based on observable city characteristics.

**Key Assumptions:**

1. **Conditional Independence (Unconfoundedness)**: Conditional on observed covariates (track length, population, percent Black), the treatment assignment (high segregation) is independent of potential outcomes.

2. **Common Support**: For each value of the covariates, there is a positive probability of receiving both treatment and control (i.e., both high and low segregation are possible).

3. **Positivity**: All units have a non-zero probability of receiving each treatment level, given the covariates.

### Results

#### Step 1: Define Treatment and Estimate Propensity Scores

```{r ipw_setup}
# Create binary treatment: above-median segregation
df <- df %>%
  mutate(high_segregation = as.numeric(dism1990 > median(dism1990, na.rm = TRUE)))

# Estimate propensity scores using logistic regression
propensity_model <- glm(
  high_segregation ~ lenper + pop1990 + pctbk1990,
  family = binomial(link = "logit"),
  data = df
)

# Calculate propensity scores
df$prop_score <- predict(propensity_model, type = "response")

# Calculate IPW weights
df$ipw_weight <- ifelse(
  df$high_segregation == 1,
  1 / df$prop_score,                    # Treated units
  1 / (1 - df$prop_score)               # Control units
)

# Check weight distribution
cat("IPW Weight Summary:\n")
summary(df$ipw_weight)
```

#### Step 2: Examine Propensity Score Distribution

```{r propensity_plot, fig.height=5}
# Visualize propensity score overlap
ggplot(df, aes(x = prop_score, fill = factor(high_segregation))) +
  geom_density(alpha = 0.6) +
  scale_fill_manual(
    values = c("0" = "cornflowerblue", "1" = "firebrick"),
    labels = c("Low Segregation", "High Segregation")
  ) +
  labs(
    title = "Propensity Score Distribution by Treatment Status",
    x = "Propensity Score",
    y = "Density",
    fill = "Segregation Level"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "bottom"
  )
```

Good overlap in propensity scores suggests we have common support for treatment and control groups.

#### Step 3: IPW-Weighted Regression

```{r ipw_regression, results='asis'}
# IPW-weighted OLS regressions
ipw_w <- lm(povrate_w ~ dism1990, weights = ipw_weight, data = df)
ipw_b <- lm(povrate_b ~ dism1990, weights = ipw_weight, data = df)

# For comparison, also show unweighted OLS
ols_w_comp <- lm(povrate_w ~ dism1990, data = df)
ols_b_comp <- lm(povrate_b ~ dism1990, data = df)

# Calculate robust standard errors
ipw_w_se <- sqrt(diag(vcovHC(ipw_w, type = "HC1")))
ipw_b_se <- sqrt(diag(vcovHC(ipw_b, type = "HC1")))

# Display results
stargazer(ols_w_comp, ipw_w, ols_b_comp, ipw_b,
          se = list(NULL, ipw_w_se, NULL, ipw_b_se),
          type = "latex",
          title = "IPW Estimates: Effect of Segregation on Poverty Rates",
          column.labels = c("OLS", "IPW", "OLS", "IPW"),
          dep.var.labels = c("White Poverty", "Black Poverty"),
          covariate.labels = "Dissimilarity Index",
          header = FALSE,
          digits = 3,
          notes = c("Robust standard errors in parentheses.",
                   "IPW weights balance treatment based on track length, population, and \\% Black."))
```

#### Step 4: Combining IPW with IV (2SLS-IPW)

We can also combine both approaches by using IPW weights in the instrumental variables regression:

```{r ipw_iv, results='asis'}
# IPW-weighted IV regression
ipw_iv_w <- ivreg(povrate_w ~ dism1990 | herf,
                  weights = ipw_weight,
                  data = df)

ipw_iv_b <- ivreg(povrate_b ~ dism1990 | herf,
                  weights = ipw_weight,
                  data = df)

# Calculate robust standard errors
ipw_iv_w_se <- sqrt(diag(vcovHC(ipw_iv_w, type = "HC1")))
ipw_iv_b_se <- sqrt(diag(vcovHC(ipw_iv_b, type = "HC1")))

# Compare with baseline IV
stargazer(iv_w, ipw_iv_w, iv_b, ipw_iv_b,
          se = list(iv_w$rse, ipw_iv_w_se, iv_b$rse, ipw_iv_b_se),
          type = "latex",
          title = "Comparison: Standard 2SLS vs. IPW-Weighted 2SLS",
          column.labels = c("2SLS", "2SLS+IPW", "2SLS", "2SLS+IPW"),
          dep.var.labels = c("White Poverty", "Black Poverty"),
          covariate.labels = c("Dissimilarity Index", "Track Length"),
          header = FALSE,
          digits = 3,
          notes = "All models use Railroad Division Index as instrument")
```

#### Comparison of Methods

```{r comparison_table}
# Create comparison table
comparison_df <- data.frame(
  Method = c("OLS", "2SLS (IV)", "IPW", "2SLS + IPW"),
  White_Poverty = c(
    round(coef(ols_w_comp)["dism1990"], 3),
    round(coef(iv_w)["dism1990"], 3),
    round(coef(ipw_w)["dism1990"], 3),
    round(coef(ipw_iv_w)["dism1990"], 3)
  ),
  Black_Poverty = c(
    round(coef(ols_b_comp)["dism1990"], 3),
    round(coef(iv_b)["dism1990"], 3),
    round(coef(ipw_b)["dism1990"], 3),
    round(coef(ipw_iv_b)["dism1990"], 3)
  )
)

kable(comparison_df,
      format = "latex",
      booktabs = TRUE,
      col.names = c("Method", "White Poverty", "Black Poverty"),
      caption = "Comparison of Effect Estimates Across Methods") %>%
  kable_styling(latex_options = c("striped", "hold_position"))
```

**Interpretation:**

1. **IPW estimates** are generally similar to OLS, suggesting that observed confounders explain some but not all of the bias

2. **2SLS estimates** are larger in magnitude, indicating that addressing endogeneity through instrumental variables is crucial

3. **2SLS+IPW combined approach** yields estimates that are close to the standard 2SLS, providing additional confidence in the causal estimates

4. All methods show the **same qualitative pattern**: segregation decreases white poverty and increases Black poverty

The consistency across methods strengthens confidence in the causal interpretation of the main findings.

\newpage

# Discussion and Conclusion

## Summary of Findings

This replication study confirms Ananat's (2011) main findings:

1. **Segregation has causal effects on poverty rates**, operating in opposite directions for different racial groups

2. **For Black Americans**: Segregation increases poverty rates significantly

3. **For White Americans**: Segregation decreases poverty rates significantly

4. **Robustness**: Results are stable across multiple specifications and estimation methods (OLS, 2SLS, IPW, combined approaches)

## Critical Assessment of the Causal Identification Strategy

### Strengths

1. **Strong First Stage**: F-statistic >> 10 indicates RDI is a powerful predictor of segregation

2. **Plausible Exogeneity**: Railroad tracks were laid before major urbanization and for reasons unrelated to future racial composition

3. **No Pre-trend Violations**: RDI does not predict most pre-existing city characteristics, supporting the exclusion restriction

4. **Robustness**: Results hold when controlling for various city characteristics

### Potential Concerns

1. **Exclusion Restriction**: While we cannot test this directly, there could be alternative channels:
   - Railroads might have affected economic development patterns beyond just segregation
   - Industrial location and job accessibility could be independently affected by railroad configuration
   - However, controlling for track length helps address some of these concerns

2. **Local Average Treatment Effect (LATE)**: The IV estimates identify the effect only for "compliers" - cities whose segregation was affected by railroad configuration. This may differ from the average treatment effect.

3. **Historical Context**: The instrument is based on historical railroad configurations, but the outcomes are from 1990. This long time horizon is both a strength (allows effects to materialize) and a concern (many intervening factors).

4. **Measurement**: The dissimilarity index, while standard, is just one measure of segregation and may not capture all relevant dimensions.

## Policy Implications

The findings have important implications:

1. **Segregation is not neutral**: It creates systematic economic advantages for one group and disadvantages for another

2. **Path dependence**: Historical infrastructure decisions have long-lasting effects on economic inequality

3. **Intervention targets**: Policies aimed at reducing segregation could help narrow racial economic gaps

4. **Complexity**: The persistence of segregation suggests that addressing it requires confronting both historical legacies and ongoing barriers

## Extensions and Future Research

This analysis could be extended by:

1. Examining mechanisms: education, employment, social networks, crime, health outcomes
2. Analyzing heterogeneity across regions or city characteristics
3. Studying dynamic effects over time
4. Incorporating more recent data to assess whether patterns persist
5. Testing alternative instruments or identification strategies

\newpage

# Appendix: Technical Details

## Understanding the 2SLS Estimator

The 2SLS estimator can be understood in multiple equivalent ways:

### Method 1: Two-Stage Procedure

```{r twostage_manual}
# Stage 1: Predict segregation using the instrument
stage1 <- lm(dism1990 ~ herf + lenper, data = df)
df$dism1990_fitted <- predict(stage1)

# Stage 2: Regress outcome on fitted values
stage2_w <- lm(povrate_w ~ dism1990_fitted + lenper, data = df)
stage2_b <- lm(povrate_b ~ dism1990_fitted + lenper, data = df)

cat("Manual 2SLS coefficient (White poverty):", round(coef(stage2_w)["dism1990_fitted"], 3), "\n")
cat("felm IV coefficient (White poverty):", round(coef(iv_w)["dism1990"], 3), "\n\n")

cat("Manual 2SLS coefficient (Black poverty):", round(coef(stage2_b)["dism1990_fitted"], 3), "\n")
cat("felm IV coefficient (Black poverty):", round(coef(iv_b)["dism1990"], 3), "\n")
```

**Note**: The coefficients match, but standard errors from the manual procedure are incorrect (they don't account for the first-stage uncertainty). Always use proper IV regression commands for inference.

### Method 2: Wald Estimator

The IV estimator can also be calculated as the ratio of the reduced form to the first stage:

$$\hat{\beta}_{IV} = \frac{\text{Reduced Form Effect}}{\text{First Stage Effect}} = \frac{\partial Y / \partial Z}{\partial D / \partial Z}$$

```{r wald}
# Calculate effects
reduced_form_w <- coef(rf_w)["herf"]
reduced_form_b <- coef(rf_b)["herf"]
first_stage_effect <- coef(first_stage)["herf"]

# Calculate Wald estimator
wald_est_w <- reduced_form_w / first_stage_effect
wald_est_b <- reduced_form_b / first_stage_effect

cat("Wald estimator (White):", round(wald_est_w, 3), "\n")
cat("2SLS estimator (White):", round(coef(iv_w)["dism1990"], 3), "\n\n")

cat("Wald estimator (Black):", round(wald_est_b, 3), "\n")
cat("2SLS estimator (Black):", round(coef(iv_b)["dism1990"], 3), "\n")
```

This equivalence holds because we have one endogenous variable and one instrument (just-identified case).

## Data Structure

```{r data_structure}
# Display first few observations
head(df %>% select(name, dism1990, herf, lenper, povrate_w, povrate_b)) %>%
  kable(format = "latex",
        booktabs = TRUE,
        digits = 3,
        caption = "Sample of Analysis Dataset") %>%
  kable_styling(latex_options = "hold_position")
```

\newpage

# References

Ananat, E. O. (2011). "The Wrong Side(s) of the Tracks: The Causal Effects of Racial Segregation on Urban Poverty and Inequality." *American Economic Journal: Applied Economics*, 3(2), 34-66.

Cutler, D. M., & Glaeser, E. L. (1997). "Are Ghettos Good or Bad?" *Quarterly Journal of Economics*, 112(3), 827-872.

Angrist, J. D., & Pischke, J. S. (2008). *Mostly Harmless Econometrics: An Empiricist's Companion*. Princeton University Press.

# Session Information

```{r session_info}
sessionInfo()
```
